prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-07-15 inst/MyArticle/paper.Rmd
update abstract.

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model quantitative proteomics experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) to model the presence or absence information of a protein, or robust linear models (_rlm_'s), in the future. R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical methods programming languages. However, the developed framework is flexible enough to in the future integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and makes selecting the best differential expression analysis algorithm for your problem easy.

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis result obtained with prolfqua are comparable to or better than when using other differential expression analysis tools.

In summary, `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. It also can generate conclusive reports and to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).
We provide documentation, in vignette format, at the website https://github.com/fgcz/prolfqua/. This document was created using Rmarkdown. All the code needed to replicate the document or the benchmark results is available at: https://github.com/wolski/prolfquabenchmark.

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. 
   The `r Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
      Finally,  The `r Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

urlcolor: blue
---

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model quantitative proteomics experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) to model the presence or absence information of a protein, or robust linear models (_rlm_'s), in the future. The developed framework is flexible enough to in the future integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. (While other R packages typically only implement one modelling approach.) Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and makes selecting the best differential expression analysis algorithm for your problem easy.

# R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical methods programming languages.

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the FDR is an unbiased estimate of the FDP. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis result obtained with prolfqua are comparable to or better than when using other differential expression analysis tools. All the code needed to replicate the benchmark results and this document is available at: https://github.com/wolski/prolfquabenchmark.

(merge with first paragraph of the Conclusions)
`r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. It also can generate conclusive reports and to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).
